Pottery decor as networks on the Middle Niger

### 2023-08-03 analysis/paper/paper.Rmd
initial commit

---
title: "Pottery decor as networks on the Middle Niger"
author:
  - Nikolas Gestrich:
      email: gestrich@uni-frankfurt.de
      institute: [FI]
      correspondence: true
  - Juan-Marco Puerta-Schardt:
      email: jmpuertaschardt@outlook.de
      institute: [GU]
      correspondence: false
institute:
  - FI: Frobenius Institute
  - GU: Goethe University
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      #pandoc_args:
      #- --lua-filter=../templates/scholarly-metadata.lua
      #- --lua-filter=../templates/author-info-blocks.lua
      #- --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

@@

Highlights: `r rmarkdown::metadata$highlights`

@@

```{r libraries, echo = FALSE}
# libraries
library(dplyr)
library(knitr)
library(patchwork)
library(ggplot2)
library(tidygraph)
library(ggraph)
library(tibble)
library(igraph)
library(tidyr)
library(sf)
```

@@

For much of the history of the archaeological discipline, ceramics have been used as markers of similarity and difference.In many areas and periods of the past, they are the most ubiquitous archaeological find category, and have thus contributed greatly to the recognition of archaeological artefact groups, or archaeological cultures. It is not rare to see find complexes and periods named after particular ceramic types. Though the current scholarly consensus is that archaeological cultures do not represent social groups as experienced by past humans, and that pots do not equal people, a large amount of work on archaeological ceramics is engaged in grouping ceramics and distinguishing groups from one another. There are a number of approaches for doing so and the rationales behind them vary. The basis for the creation of archaeological cultures, especially current in works of the first two thirds of the twentieth century, is the 'traditional' typological approach to ceramics and other artefacts. This approach is nowadays most used in the best-explored regions with the longest research traditions, and groups ceramics mainly according to outward appearances, such as shape and surface treatment. The distribution, both spatially and temporally, of such types then becomes the subject of research, which subsequent studies can augment through the recognition of sub-types and variants. A different approach takes as its basis the sequence of manufacture of artefacts, the *chaîne opératoire*. This approach proposes that the manufacture of an artefact involves a series of choices constrained by certain outside factors. These choices are seen as being cultural, so that a similar set of choices under similar constraints are markers of shared cultural knowledge. Here, the archaeological research goal also lies in the identification of spheres of cultural commonality which can be distinguished from one another.

In the archaeology of West Africa, and especially in the wider areas surrounding the upper and middle reaches of the Niger, there has been a strong research focus on ceramics. From the 1st millennium BC onwards, they occur in extremely high numbers on archaeological sites, and their appearance, especially their surface treatment, is highly variable. In contrast to the Great Lakes region of Eastern Africa, for example, the number of different roulette decorations, painted, incised, and applied decors, but also the variety in form, is very high. Consequently, there has been considerable emphasis on ceramics in archaeological and also in ethnoarchaeological work. While much of the ethnoarchaeological work has been based on exploring *chaînes opèratoires*, the archaeological work has mainly relied on attribute-based analyses (the system employed by S.K. McIntosh at Jenné-jeno has been very influential in subsequent work). In a region where large archaeological sites abound and researchers are relatively thin on the ground, the main hope of the analyses usually is to create chrono-stratigraphical sequences in order to be able to give approximate periods to surface assemblages. Despite all the variability, it has been found that a great number of ceramic attributes are not particularly time-sensitive. On the contrary, most sites appear to have diverse assemblages in which the percentages of individual attributes vary, but persist over several centuries, sometimes millennia.

Some researchers have used the attribute-based approach to define types, statistically derived for the most part, which typify the ceramics of a certain place of time (MacDonald's Facies, McIntosh's timeline). The spatiotemporal distribution of individual attributes has also been the subject of some important work, such as individual rouletted decors (Livingstone-Smith) or base-shaping techniques (Huysecom). The most comprehensive treatment so far is in the work of Anne Mayor, combining both shaping techniques and decorative practices from an archaeological and ethnoarchaeological point of view. Focused on the Bandiagara highlands in central Mali, Mayor uses her data to construct an impressive overview of the development of different techniques and the establishment of ceramic traditions, which she closely relates to ethno-linguistic groups. Mayor's work builds on a large body of data and decades of ethnoarchaeological fieldwork, and there is little to add to her conclusions. Here, however, we wish to show that an almost identical dataset as available to Mayor can generate quite a different perspective when analysed as a network, and when connectivity is emphasised over boundaries.

# Archaeological ceramics as networks

In recent years, the use of network analysis in archaeology has been growing significantly. A large number of these studies focus on ceramics, but they are geographically mainly limited to North America (e.g. Hart 2012, Chiu 2019, Barker and Young 2017, Crabtree 2015, Mills et al. 2013, Peeples and Haas 2013) and the Mediterranean (e.g. Sternini 2019, Iacono 2016, Brughmans and Poblome 2016 torrence & Swadling 2008, de Groot 2019). The archaeological records of these regions are amongst the best studied in the world, with fine-grained pottery typologies and extensive excavations. To our knowledge, this is the first application of a network approach to archaeological material from sub-Saharan Africa, where the conditions are somewhat different.

Ethnographic studies of pottery manufacturing in Africa suggest that the way that pottery is made depends largely on systems of learning and teaching the craft (Gosselain), on local market consumer choices (Dietler & Herbich), or on marriage patterns amongst manufacturing groups (ibid.). In short, the processes underlying the propagation of technical practices through time and across geographical space are dependent on communication and, more specifically, knowledge exchange. Such exchange occurs between humans and we can therefore understand similarities in the *chaine opératoire* as evidence that knowledge was shared. However, different elements of the pottery making process are thought to be reflective of different kinds of social interaction. Gosselain (REF), whose research in the years between XX and XX explored this question, concludes that, on the whole, forming techniques are more stable, reflective of durable cultural ties, sometimes even ethnic groups. On the other end of the spectrum, decorations are thought to be influenced by far more ephemeral contact, in which seeing someone else's pot might be enough to change a potter's production, as could indirect contact via the wishes of customers.

SNA shifts the focus "from spatial relations to social relations" (cf. Knappett 2011)
Mills et al. (2015) have been able to compare similarity in wares across sites, but this cannot be done here. They also highlight Knappet's (2011) idea of different scales in archaeological networks. Because of the specific use case scenario here, we are conflating two fo his scales. He sees the community of practice as a meso-scale effect, across maybe a few households. However, we can see in the archaeological ceramic data from Mali that communities of practice in individual stages of the *chaîne operatoire* is often much larger. Thus, certain parts of the pottery manufacturing sequence are distributed across larger or smaller networks, but rarely remain at the household level. Also, we do not really have the luxury of being able to pinpoint certain wares, since typology is not as established.

We need to be aware of the fact that we are conflating consumption with production here. This is why we will need to threshold the networks, to make sure that a site represents the significant potting communities within its area.

@@

```{r import 11c decor data, echo = FALSE}
library(readr)
# import decor data
decors <- read_csv2(here::here("analysis/data/raw_data/decors_late.csv"))
# import information on sites
sites <- read_csv2(here::here("analysis/data/raw_data/sites_all.csv"))
```

```{r datatable, fig.cap="decor data used for network construction"}
library(kableExtra)
kable(decors, digits = 1, "html") %>% 
  kable_styling("condensed") %>% 
  save_kable(here::here("analysis/figures/datatable.html"))
```

In this paper, we use pottery decor data from contexts occupied in the 11th century AD to attempt a reconstruction of networks through which pottery decorative practices were shared. The data is in a common format of percentages of prevalence in the assemblage (see \@ref(tab:datatable) This particular time slice, which represents a cumulative view of a great many individual interactions, will then be compared to earlier contexts in the same wider area, in order to demonstrate the diachronic possibilities of the approach.The sites we have been able to include in the network model are shown in \@ref(fig:sitemap). We can see here that sites are geographically unequally distributed. This is due to research history as well as past cultural dynamics. Certainly, the IND and the areas bordering it (notably the Lakes region) were strongly populated at the time, maybe more so than other areas. The cluster of sites around the Bandiagara escarpment is due to the extraordinarily detailed and long-term research of the University of Geneva's research teams. Blank spaces do not imply a lack of people, in this case people making and using pottery. Global politics and research priorities over the last 50 years have meant that these are areas we do not currently have excavated and dated ceramics for, though we know that sites exist, often in abundance. This means that, when we construct networks, we do so with a geographically unrepresentative sample, which leads to incomplete networks. Sindbaek has pointed out the difficulty in this situation: the actual network is not known, only its effects, from which the network can be partially reconstructed. Accordingly, this study should be seen as an exercise first in the reconstruction of a possible network that led to the data preserved in the archaeological record: a tiny sample of a tiny sample, influenced by an unknown amount of mostly unknowable pre- and post-depositional biases. Nevertheless, we will attempt to analyse the reconstructed network in order to formulate some hypotheses about the nature of innovation and spread in ceramic decors. Future generations of researchers are encouraged to improve upon this initial attempt.

```{r sitemap, fig.cap="location of sites in the analysis", echo = FALSE}
source(here::here("analysis/R/map plots.R"))

sitepoints <- st_as_sf(sites, coords = c("lon", "lat"), crs = 4326, agr = "constant")

sitemap <- ggplot()+
  map_bw_grid +
  geom_sf(data = sitepoints)

sitemap + inset_element(inset, 0.72, 0.72, 1, 1)
```

The network is constructed from standard archaeological frequency data, in which the proportions of decorative practices are represented across the selected site phases (REF FIG XXX). The decorative practices encode both the tool used and the action performed, for instance a twisted cord (tool) roulette (action). This data is transformed into a network by deciding that every decor that is shared between two sites leads to a link between these sites. The link is weighted by multiplying the proportion at one site with the proportion at the other. This leads to a network in which the nodes (sites) can be connected in more than one manner, known as a multiplex or multilayer network. As is shown in \@ref(fig:total-network), this leads to a very tightly connected network, known as a "hairball". However, the multiplex nature of the network allows us to make an important initial observations on its structure: the networks in which individual decor types are propagated are overlapping. There are no two decor networks that cover one another entirely. This emerges more clearly in \@ref(fig:total-network-sample), when the number of decorative practices is reduced. This indicates that the adoption of decorative techniques ...

@@

# construct network, remove extraneous sites
decor_graph <- tbl_graph(sites, edgelist, directed = FALSE) %>%
  activate(nodes) %>%
  filter(!node_is_isolated())
```

```{r total-network, fig.cap="Plot of 11th century AD network of decorative practices. The plot shows the very strongly connected structure of the network."}
ggraph(decor_graph, layout = "stress") +
  coloured_edges +
  theme_graph()
```

```{r total-network-sample, fig.cap="Partially overlapping networks of four decorative practices", echo = FALSE}
decor_graph %>%
  activate(edges) %>%
  filter(name %in% c("braided_cord_roulette", "braided_strip_roulette", "paint", "straight_cord_mat")) %>%
  ggraph(layout = "stress") +
  coloured_edges +
  theme_graph()
```

Using the `multinet` package, the statistical distance (Jaccard similarity) between the different decors' networks were calculated based on the nodes they contained. We see in \@ref(fig:layer-comparison) that some networks are more similar to one another than others. This leads us to ask whether there might be groups of decors that were shared within similar networks, and thus, whether there might be groups of sites predominantly sharing specific decors.

```{r import to multinet, echo = FALSE}
library(multinet)
# reduce graph to types with over 5 edges
good_edges <- decor_graph %>%
  activate(edges) %>%
  as_tibble() %>%
  count(name) %>%
  filter(n > 5)

reduced_graph <- decor_graph %>%
  activate(edges) %>%
  filter(name %in% good_edges$name)

# prepare graph and make a vector of different layer names
prep.graph <- reduced_graph %>%
  activate(edges) %>%
  rename(layer = name) %>%
  activate(nodes) %>%
  rename(name = label)

layers <- prep.graph %>%
  activate(edges) %>%
  as_tibble() %>%
  distinct(layer) %>%
  pull(layer)

# get a list of individual networks using the onelayer function (/R/network functions.R)
graphlist <- purrr::map(layers, onelayer)
names(graphlist) <- layers

>>>>

---
title: "Pottery decor as networks on the Middle Niger"
author:
  - Nikolas Gestrich:
      email: gestrich@uni-frankfurt.de
      institute: [FI]
      correspondence: true
  - Juan-Marco Puerta-Schardt:
      email: Puerta-Schardt@em.uni-frankfurt.de
      institute: [GU]
      correspondence: false
institute:
  - FI: Frobenius Institute
  - GU: Goethe University
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      #pandoc_args:
      #- --lua-filter=../templates/scholarly-metadata.lua
      #- --lua-filter=../templates/author-info-blocks.lua
      #- --lua-filter=../templates/pagebreak.lua
abstract: |
  This paper reports on an attempt to analyse decorative techniques on archaeological pottery from the Middle Niger as knowledge exchange networks. The 12th-century CE state of this network is modelled and analysed, and then compared to its 9th-century counterpart. The analysis shows how knowledge about pottery decoration was propagated within networks of different sizes and intensities, from local, well-connected networks to more loosely tied long-distance connections. The article shows how the perspective of network analysis differs from, and can fruitfully complement, previous treatments of ceramic décor distribution in the region. It adds a novel perspective to what the distribution of archaeological ceramics in this region reflects, and contributes to the generation of hypotheses that can be further tested by fieldwork. 
keywords: |
  Mali; network analysis; pottery; archaeological ceramics
---

@@

@@

```{r libraries, echo = FALSE}
# libraries

# Load the dplyr package for data manipulation
library(dplyr)

# Load the knitr package for generating reports using RMarkdown
library(knitr)

# Load the patchwork package for combining multiple plots into a single plot
library(patchwork)

# Load the ggplot2 package for creating graphics and plots
library(ggplot2)

# Load the tidygraph package for working with tidygraph objects
library(tidygraph)

# Load the ggraph package for visualizing graph data
library(ggraph)

# Load the tibble package for creating and manipulating tabular data
library(tibble)

# Load the igraph package for working with network data
library(igraph)

# Load the tidyr package for data tidying and reshaping
library(tidyr)

# Load the sf package for reading, writing, and manipulating spatial data
library(sf)

# Load the multinet package for analysis of multilayer networks
library(multinet)

# Load the kableExtra package for nicer table design
library(kableExtra)

# Load the readr package to read excel-data
library(readr)
```

@@

The 40 years between 1970 and 2010 were a period of increasing archaeological research in Mali, West Africa. In the region around the middle reaches of the river Niger, this research documented the remains of a dynamic civilization which, from the last millennium BC onwards, developed urbanism, craft specialization, and long-distance trade (McIntosh 1998), and was highly influential not only in changing narratives about the African past but also in global debates on sociocultural dynamics (McIntosh 1999). One of the characteristics of many settlement sites of the first-early second millennium AD in the region is the high volume of ceramic sherds, which sometimes challenged the capacities of research projects (McIntosh 1995, pp. 132--133). Featuring a large variety of decorative motifs, forms, and production techniques, the pottery of the region has been the subject of numerous studies, both from an archaeological and ethnoarchaeological perspective. Three key points have emerged from this research history. Firstly, establishing chronologically sensitive typologies has been very difficult, with contemporary variability often more significant than diachronic change. Secondly, base forming techniques, which vary throughout the region, are likely to correspond to larger cultural and maybe also linguistic spheres. Thirdly, pottery vessels appear to have been objects of trade, at times over long distances.

In this paper, we consider decorative practices on 12^th^ century AD ceramics of this region. We use a network approach to ask how such practices might have spread, and, even when they are not useful as chronological markers, how they might still inform us about sociocultural dynamics of the ancient Middle Niger. This research aim reflects a long-standing concern of archaeologists in the area, but the network approach we adopt here shifts the perspective from ceramic production as representing sociocultural entities to the processes by which techniques spread. This results in connections, rather than boundaries, as the subject of research. Using network models, we ask what ceramics can tell us about the processes that integrated the Middle Niger as a cultural area despite its diversity, and where cultural exchange might have been facilitated or impeded, and why.

## Middle Niger Ceramics:

In many contemporary societies in the region, pottery is made by occupational specialist groups (Bambara: *nyamakala*). Though there are exceptions, it is the cultural norm in the wider Middle Niger region that pottery is made by women of groups in which the men are specialists in metallurgy. There is some debate as to how far in the past this specialization developed, but it currently seems likely that it has its roots in the first millennium AD (McIntosh 1998, pp. 177--178), and that it had spread throughout the region by the mid second millennium (Tamari 1991, 1995). There is no doubt that potting and metallurgical experts were of great significance to the economy of the Middle Niger during this period of urbanism and emerging statehood, since they made basic products used in most parts of everyday life, from food production to the religious sphere. The cultural innovations linked to pottery making, both in social terms (endogamous craft specialists) and in technology (pottery and metal products), are a key factor in the development of the Middle Niger civilization (McIntosh 1998).

From the 1st millennium BC onwards, ceramics occur in high numbers on archaeological sites on the Middle Niger, and their appearance, especially their surface treatment, is highly variable. The number of different roulette decorations, painted, incised, and applied decors, but also the variety in form, is very high, especially when attention is paid to the techniques and tools of decoration, rather than exclusively to their effects. Consequently, there has been considerable emphasis on ceramics in archaeological and in ethnoarchaeological work. Early studies (Bedaux 1980; e.g. Curdy 1982; and particularly Bedaux and Lange 1983) noted both regional differences in the appearance of pottery as well as striking widespread similarities in individual forms or attributes. Subsequent work has seen both in-depth studies on ceramics of individual sites (e.g. McIntosh 1995; Schmidt et al. 2005; Gallin 2011; van Doosselaere 2014) as well as studies on the spread of particular decorative practices (Livingston Smith 2007) or shaping techniques (e.g. Sterner and David 2003; Mayor 2011a). Three main features of the wider region's ceramic record are important to this study: 1) the relative lack of clear temporal change, 2) high contemporary variability, and 3) pottery as exchange items.

### Lack of temporal markers:

Susan K. McIntosh's analyses of the ceramics of Jenné-jeno and surrounding sites resulted in a first detailed pottery chronology on the Middle Niger. In two studies (McIntosh and McIntosh 1980; McIntosh 1995), McIntosh was able to show how some pottery shapes, fabrics and decorations showed clear temporal trends. There was, however, also quite strong stability, particularly in many of the decorative techniques (McIntosh and McIntosh 1980, pp. 233--244). Subsequent studies on other sites have also shown change in ceramic production, and particularly in its decorative aspects, to have been very gradual, and stability to have been the norm (e.g. Schmidt et al. 2005). This contrasts with the archaeological record in many other regions of the world, where the outward appearance of pottery has been far more changeable over time and can therefore be used as chronological markers. This has not been possible for most periods and areas of the Middle Niger region, where decorative aspects tend only to vary in their proportion through time.

### Ceramics and social boundaries:

The more striking differences between archaeological ceramics in the area are not diachronic, but contemporary. Pottery from different areas varies in its forming techniques, its shapes (though the most pottery is globular or ovoid in its basic shape), and in decorative practices. Occasionally, as at Tongo Maaré Diabal (Gestrich and MacDonald 2018) or Kumbi Saleh (van Doosselaere 2014), different productions can be clearly separated. The mechanisms underlying these differences have been the subject of extensive ethnoarchaeological and archaeological research by the University of Geneva. Led from the outset by the goal of relating ceramic production to ethnic groups (Gallay et al. 1998, pp. 15--22), this extensive research documented how certain parts of the ceramic production -- notably base forming techniques and tools employed in the potting process -- could be employed to create typologies that were patterned by the ethnic identity, economic milieu, and ecological zone of the producers (Gallay et al. 1998, pp. 106--107; Mayor 2011b, p. 14). But certain decorative techniques were also found to have a certain correlation to the ethnicity-based typology created by the Geneva team (Mayor 2011b, pp. 40--78). The main limiting factor here, as Mayor stresses, is that "stylistic borrowings are relatively easy and frequent" (Mayor 2011b, p. 78), but given the fact that some of the decors do correlate with ethnic groups --- or at least with the types assigned to ethnic groups --- Mayor hints at the suggestion that decorative techniques can historically be assigned to one or another group from which they originated (ibid.).

Based on their extensive fieldwork, Mayor's hypothesis is that pottery making is learnt within stable and relatively impermeable family relationships. This is in keeping with some of Gosselain's results from Niger, where he has suggested that this initial approach to the shaping of a pot was the most influenced by teaching and learning (Gosselain 2000, p. 193). Since this usually happens from an early age within the family, it makes sense that kinship relationships, and thus often group and linguistic belonging, would influence the propagation of such techniques. These processes of knowledge exchange, limited to endogamous groups and ethnic milieus, are what explains the commonalities that underlie the typologies created by the Geneva group. Exchange outside of these limits is seen as the exception, rather than the norm. Nevertheless, Mayor gives the following scenarios for the diffusion of pottery traits (Mayor 2011b, p. 83): Firstly, when a technique is simple and expedient, thus easy and profitable to learn. Secondly, when an ethnic group is in the minority in a certain area or when communities find themselves at the geographical margins of their ethnic group. Thirdly, rare cases in which the social order is transgressed by marriage outside of the ethnic group. Mayor thus expects that shaping and decorative techniques would have had a single and ethnically distinct origin and that it is "several centuries of borrowing" (ibid.) that have led to the less demarcated situation that the Geneva team observed in the 1990s and 2000s.

### Ceramics as trade goods:

Even though most ceramic production appears to have been made for local markets, there is also evidence of pottery being used in trade over longer distances. Historically, the practice of pottery being moved on the Niger by boat and sold widely throughout the region has been documented in an account from the 18^th^ century (Park 1799, p. 219). From the archaeological evidence, it seems as though this was not confined to later periods, but at least occasionally practiced also in the early second millennium and before. Finds of imported ceramics have, for instance, been made at Essouk (Nixon 2017, p. 143,146-147), and at Kumbi Saleh (van Doosselaere 2014, pp. 228--233). We must therefore always consider the possibility that archaeological assemblages do not only reflect local production, and add the indirect contact from long-distance trade to the possible mechanisms of change in ceramic practices.

If the goal of the work by Gallay and Mayor and their teams was to identify the borders between groups in the past, this study asks a different question: how was pottery knowledge transmitted? What were the factors limiting or encouraging the flow of information? We thus begin from a null hypothesis of unlimited sharing, rather than of delimited groups, to see how the results might complement those of the Geneva research teams. Using pottery décor as a relatively easily borrowed attribute, we aim to generate hypotheses on how pottery making knowledge spread along the ancient Middle Niger. Further, we wish to demonstrate the use of network analyses for approaching West African archaeological ceramics and for making such connections visible and explorable. This technique presents a change of perspective on the ceramics of this region, allowing questions of connection to take precedence over those of boundaries.

## Ceramic decors as networks

In recent years, the use of network analysis in archaeology has been growing significantly (see e.g. Knappett 2021). A large number of these studies focus on ceramics, but geographically most important are North America (Hart 2012; Mills et al. 2013; Crabtree 2015; e.g. Barker and Young 2017; Vavrasek 2020) and the Mediterranean (Torrence and Swadling 2008; Brughmans and Poblome 2016; Iacono 2016; de Groot 2019; Sternini 2019). The archaeological records of these regions are amongst the best and longest studied in the world, with fine-grained pottery typochronologies and extensive excavations. To our knowledge, this is the first application of a network approach to archaeological ceramic material from sub-Saharan Africa, where the conditions are somewhat different (McIntosh 1995, pp. 130--131). That is not to say that archaeological pottery on the Middle Niger is poorly studied. Most researchers in the wider area have detailed publications on their ceramic assemblages, efforts have been made at systematizing nomenclature (Haour et al. 2010) and at synthetic overviews (Mayor 2011b). A large body of ethnographic and ethnoarchaeological literature provides the theoretical basis that informs the network model.

Ethnographic studies of pottery manufacturing in Africa suggest that the way that pottery is made depends largely on systems of learning and teaching the craft (Gosselain 1998), on local market consumer choices (Dietler and Herbich 1994), or on marriage patterns among manufacturing groups (ibid.). In short, the processes underlying the propagation of technical practices through time and across geographical space are dependent on communication and, more specifically, knowledge exchange. Such exchange occurs between humans and we can therefore understand similarities in the *production* as evidence that knowledge was shared. However, different elements of the pottery making process are thought to be reflective of different kinds of social interaction. Gosselain (2000) concludes that, on the whole, forming techniques are more stable, reflective of durable cultural ties, sometimes even ethnic groups. On the other end of the spectrum, decorations are thought to be influenced by far more ephemeral contact, in which seeing someone else's pot might be enough to change a potter's production, as could indirect contact via the wishes of customers.

Decoration on pottery vessels is a topic that has been central to archaeological debates for a very long time. To construct an overly simplified history of ceramic decors in African archaeology, we might distinguish an early period analogous to other parts of the world, in which decors (as a main constituent of typologies) served to distinguish human groups. This position became more elaborate and changed rationale over time, arguably culminating in Hodder's "Symbols in Action" (1982). David et al.'s "Why pots are decorated" (1988) did much to remove confidence in this idea and to situate decorative practices as part of chains of production and expressions of often unconscious propagation of social values. For them, "designs on pottery, far from being 'mere decoration,' art for art's sake, or messages consciously emblemic of ethnicity, are low-technology channels through which society implants its values in the individual --- every day at mealtimes" (David et al. 1988, p. 379). Over the intervening years, the comparatively strong ethnoarchaeological research in Africa has developed an increasingly complex picture of ceramic décor. Gosselain's (2011) "pourquoi le decorer", part review, part programmatic article, gives a good overview : Decorations on pottery occur within cultural constraints and logics, are expressions of individuality, are market-oriented, have practical considerations, are made for visual effect, have profound meanings while being thought of as trivial, are subject to fashion and taste, stand for group membership, mean different things to different people. In short, pottery decoration is not anchored in any stable way to any cultural phenomenon. More importantly for this study, however, Gosselain points out that the propagation, the process of spreading, of decorative practices is a more fruitful topic for study (ibid.: 12). Various decorative practices presuppose various kinds of contact. Some of these, as in the case of a particularly intricate rouletted tool, require the sharing of cultural knowledge and probably reflect shared practices in other cultural domains as well. Others are transmitted more easily, by sight alone. Whatever these mechanisms, there are constraints on transmission and innovation. As Herbich (1987, p. 202) notes, potters innovate and express themselves with reference to other potters they interact with. Therefore, the innovation and spread of decorative practices does not occur freely, but is helped by geographical proximity, social contact, and cultural factors. It is likewise hindered by geographic distance and cultural differences.

This, then, is the basis on which our social network model is conceived. It supposes that co-occurrence of a given practice of decorating pottery is emblematic of a form of sustained social contacts, within the limitations mentioned above. These contacts are not as close as we would expect from a teaching and learning environment, which would be better reflected in forming techniques, but are sustained and meaningful enough for there to have been an exchange of cultural aesthetic values and an exchange of knowledge of how to decorate pottery. These exchanges are therefore what the ties in our network represent.

In choosing sites as the nodes of the network, the data come mainly from contexts in which pottery was consumed, rather than produced. The network agents in the model are therefore communities of consumers. Their ties, based on shared decors on their pottery, are indirect: they must be understood to have been connected via producers (individuals and communities) and possibly also merchant middlemen which remain largely unseen in the archaeological record. Since we have established at least occasional long-distance trade in pottery in our research area (see above), this is clearly a factor limiting the model's explanatory power. Nevertheless, we can use the work carried out by the Geneva MAESAO project to suggest that the majority of pottery found at a site will generally be locally produced (Mayor 2011b, p. 84), which might reasonably allow some conflation between pottery producers and consumers in the network nodes. A low level of imported ceramics can be adjusted for in the network model, but should not be excluded, as these "weak ties" have been shown to be important for introducing innovation in social networks (Granovetter 1973), including technical innovation in ceramic production (Roux et al. 2018).

The analysis of the network will attempt to generate hypotheses on how such knowledge exchanges were structured, the level of interaction between inferred producers of ceramics, and the overall level and structure of cultural contact in the region. This differs somewhat from the rationales for pottery décor networks in studies from other parts of the world. Hart and Engelbrecht (2012, pp. 324--330) justify their building of networks based on pottery décor similarity by arguing for décors as signals in which both group membership and individual skill are expressed. Iacono (2016), on the other hand, takes the co-occurrence of pottery decoration as analogous to other forms of social encounters, and thus as representative of them, conveying the agency of individuals "simply through sensory recognition" (Iacono 2016, p. 126). We would like to stress here that these differences are in no small part due to the history of research and the nature of the archaeological record in different parts of the world, and that the building of network models on archaeological data always requires a well-theorised foundation appropriate to the historical context of the studied material.

@@

```{r import 11c decor data, echo = FALSE}
library(readr)
# import decor data

decors <- read_csv2(here::here("analysis/data/raw_data/decors_late.csv")) |>
  select(-1)

# import information on sites
sites <- read_csv2(here::here("analysis/data/raw_data/sites_all.csv"))
```

## Data

In this paper, we use pottery decor data from contexts occupied in the 12th century AD to attempt a reconstruction of networks through which pottery decorative practices were shared. The network is constructed from standard archaeological frequency data, in which the proportions of decorative practices are represented across the selected site phases (the raw data are available as Annex 1). The decorative practices encode both the tool used and the action performed, for instance a twisted cord (tool) roulette (action). This data is partly taken from existing literature, partly relies on our own fieldwork (TMD and TOG2) or raw data raised by other researchers (TBZ), and was partly documented in collections (Gao-Saney, Tellem). This time slice, which represents a cumulative view of a great many individual interactions, will then be compared to earlier contexts in the same wider area, in order to demonstrate the diachronic possibilities of the approach. The sites we have been able to include in the network model are shown in \@ref(fig:sitemap). We can see here that sites are geographically unequally distributed. This is due to research history as well as past cultural dynamics. Certainly, the IND and the areas bordering it (notably the Lakes region) were strongly populated at the time, maybe more so than other areas. The cluster of sites around the Bandiagara escarpment is due to the extraordinarily detailed and long-term research of the University of Geneva's research teams. Blank spaces do not imply a lack of people, in this case people making and using pottery. Global politics and research priorities over the last 50 years have meant that these are areas we do not currently have excavated and dated ceramics for, though we know that sites exist, often in abundance. This means that in addition to the inherent difficulty of archaeological network models -the actual network is not known, only its effects, from which the network can be partially reconstructed (Sindbaek (2013)-, the effects of missing areas or sites on the network need to be calculated (Peeples et al. 2016).Accordingly, this study should be understood as an exercise in the reconstruction of a possible network that led to the data preserved in the archaeological record: a tiny, unrepresentative sample of a tiny unrepresentative sample, influenced by an unknown amount of mostly unknowable pre- and post-depositional biases. Nevertheless, we will attempt to analyse the reconstructed network to formulate some hypotheses about the nature of innovation and spread in ceramic decors. Future generations of researchers are encouraged to improve upon this initial attempt.

```{r sitemap, fig.cap="location of sites in the analysis", echo = FALSE}
source(here::here("analysis/R/map plots.R")) # Load the required R script

#saving site location as sf
sitepoints <- st_as_sf(sites, coords = c("lon", "lat"),
                       crs = 4326, agr = "constant")

#creating a map of mali with the sites
sitemap <- ggplot() +
  map_bw_grid +
  geom_sf(data = sitepoints)

#ploting map with inset
sitemap + inset_element(inset, 0.72, 0.72, 1, 1)
```

## Network construction

Our theoretical approach to what the network represents informs the way in which the network model is constructed. So far, the most frequent form of making ties in archaeological ceramic data has been by calculating the statistical similarity between assemblages. (Hart and Engelbrecht 2012) use the Brainerd-Robinson similarity coefficient, as do, for instance, (Mills et al. 2015), while (de Groot 2019) uses the Jaccard and Kulczyinski-2 similarity measures. The similarity value is then used as the basis and weight for forming a tie between the two assemblage nodes. Since we did not only want to base our network on overall similarity but wanted to be able to closer analyze the networks of individual decors, we have constructed a multiplex network. This is a network in which edges can have different types, and thus the network can be envisaged as having layers. A multiplex network is distinct from a multilayer network, in which connections can also exist between layers, and which has recently seen its first application to archaeological ceramics by (Upton 2019). In our multiplex network, each decorative practice is represented as a distinct layer. The layers share the site assemblages as nodes, though not every node is necessarily represented on every layer. The ties are weighted by the product of the percentages of a given décor in two assemblages. This network construction is implemented in R (R Core Team 2022) using the tidygraph package (Pedersen 2022).

# Analysis

## Multiplex structure

As is shown in \@ref(fig:total-network), the initial structure of the data leads to a very tightly connected network, known as a "hairball". This is normal for archaeological networks based on co-occurrence or similarity, since even rare occurrences lead to a tie. However, the multiplex nature of the network allows us to make an important initial observations on its structure: the networks in which individual decor types are propagated are only partly overlapping. There are no two decor networks that cover one another entirely. This pattern emerges more clearly in \@ref(fig:total-network-sample), when the number of decorative practices represented on the graph is reduced. This structure suggests that the networks allowing decorative practices to propagate were at least partially different in every case. Remembering the constraints on spread identified by Gosselain (2011, see section 2), we might propose that external factors, such as geography and cultural differences, provided "frontiers" to the spread, albeit different ones in each case. Some decors are very widespread, and thus had fewer inhibitions to their spread, while others occur less frequently.  We can now proceed to further analyse the network in order to identify where such frontiers and catalysts of spread might have been and how they were structured.

@@

# construct network, remove extraneous sites
decor_graph <- tbl_graph(sites, edgelist, directed = FALSE) |>
  activate(nodes) |>
  filter(!node_is_isolated())
```

```{r total-network, fig.cap="Plot of 11th century AD network of decorative practices. The plot shows the very strongly connected structure of the network."}
# Creating a multi-layer graph
ggraph(decor_graph, layout = "stress") +
  coloured_edges +
  theme_graph()
```

```{r total-network-sample, fig.cap="Partially overlapping networks of four decorative practices", echo = FALSE}
decor_graph |>
  activate(edges) |>
  # Filter "edges" based on 4 specific decorations
  filter(name %in% c("braided_cord_roulette",
                     "braided_strip_roulette",
                     "paint",
                     "straight_cord_mat")) |>

  #plot the graph
  ggraph(layout = "stress") +
  coloured_edges +
  theme_graph()
```

In the first place, we will determine whether there are decors that share similar knowledge exchange networks and others that are dissimilar. Using the multinet package (Magnani et al. 2021), the statistical distance (Jaccard similarity) between the different decors' networks were calculated based on the nodes they contained.

\@ref(fig:layer-comparison) shows that some networks are more similar to one another than others. For instance, the practice of rouletting with a braided cord  (McIntosh and Gueye 2010) appears to have a significant shared network with that of rouletting with a folded strip (Haour and Keita 2010). This means that knowledge of the two tools necessary for these roulettes (for details, see Haour et al. 2010) spread in a similar geographic and cultural sphere, without being impeded. On the other hand, there appears to have been quite strong boundaries in place between the networks sharing knowledge on rouletting with a braided strip and dragging a comb along the surface of the pot or using painted decoration. There thus appear to have been groups of decors that were shared within similar networks, and we can thus ask whether there might have been communities in the network predominantly sharing specific decors.

```{r import to multinet, echo = FALSE}
library(multinet)
# reduce graph to types with over 5 edges
good_edges <- decor_graph |>
  activate(edges) |>
  as_tibble() |>
  count(name) |>
  filter(n > 5)

#creating a reduced graph with only those edges
reduced_graph <- decor_graph |>
  activate(edges) |>
  filter(name %in% good_edges$name)

# prepare graph for multinet by renaming
prep_graph <- reduced_graph |>
  activate(edges) |>
  rename(layer = name) |>
  activate(nodes) |>
  rename(name = label)

#make a vector of different layer names
layers <- prep_graph |>
  activate(edges) |>
  as_tibble() |>
  distinct(layer) |>
  pull(layer)
